---
title: PA Socioeconomic Conditions
author: "Dr. Alexis R. Santos"
date: "March 9, 2018"
output:
  word_document: default
  pdf_document: default
  html_document: default
always_allow_html: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Overview
This code will produce socioeconomic conditions for PA. 

#Accessing county level data for empirical comparisons
In this section, we access demographic variables required to replicate the regression analyses presented in the article by Lichter, D. T. and McLaughlin, D. K. (1995). Data in the ACS will allow us to obtain the following variables:

* Demographic Compositon with ACS table
  + Percent black: B03002
  + Percent Latino (Hispanic): B03002
  + Percent aged less than 18: B01001
  + Percent aged 65 or older:  B01001
  + Percent immobile: B07003 now this is measured as one-year, not 5-years. 
  + Percent less than High School Diploma: B15003
  + South: Region
  
* Industrial Structure with ACS table 
  + Percent manufacturing: C24050
  + Percent extractive: C24050
  + Percent service: C24050
  + Percent government (Public Administration): C24050

* Employment with ACS table in parenthesis
  + Percent unemployed: B12006
  + Percent females employed: B12006

* Family Structure with ACS table in parenthesis
  + Percent female headed households: B25011.
  
The following lines of code create the variables we will request using the "get_acs" function. Careful attention must be placed on the number of individuals in our group of interest and in the denominator for the rates we are calculating. Please, review the variables we will get from the Census API and corresponding denominators (_001E). 

```{r create the object for variables}

census_api_key("Your Census Key Here",install=TRUE)

paper_variables<-c("B03002_004E","B03002_012E","B03002_001E","B01001_003E",
                   "B01001_004E","B01001_005E","B01001_006E","B01001_027E",
                   "B01001_028E","B01001_029E","B01001_030E","B01001_001E",
                   "B01001_020E","B01001_021E","B01001_022E","B01001_023E",
                   "B01001_024E","B01001_025E","B01001_044E","B01001_045E",
                   "B01001_046E","B01001_047E","B01001_048E","B01001_049E",
                   "B15003_002E","B15003_003E","B15003_004E","B15003_005E",
                   "B15003_006E","B15003_007E","B15003_008E","B15003_009E",
                   "B15003_010E","B15003_011E","B15003_012E","B15003_013E",
                   "B15003_014E","B15003_015E","B15003_016E","B15003_001E",
                   "C24050_004E","C24050_002E","C24050_029E","C24050_001E",
                   "C24050_014E","B12006_006E","B12006_011E","B12006_017E",
                   "B12006_022E","B12006_029E","B12006_033E","B12006_039E",
                   "B12006_044E","B12006_050E","B12006_055E","B12006_004E",
                   "B12006_009E","B12006_015E","B12006_020E","B12006_026E",
                   "B12006_031E","B12006_037E","B12006_042E","B12006_048E",
                   "B12006_053E","B12006_010E","B12006_008E","B25011_013E",
                   "B25011_037E","B25011_001E","B07003_004E","B07003_007E",
                   "B07003_001E","B17001_002E","B17001_001E","B17005_006E",
                   "B17005_011E","B17005_017E","B17005_022E","B17005_001E",
                   "B12006_021E","B12006_032E","B12006_043E","B12006_054E",
                   "B12006_019E","B12006_030E","B12006_041E","B12006_052E","B19083_001E","B25099_001E",
                   "B25077_001E","B27010_001E","B27010_017E","B27010_033E","B27010_050E","B27010_066E")

#library(acs)
library(tidycensus)
library(tidyverse)
library(tigris)
library(sf)

#Get ACS data for county level, PA, and the 2015 ACS release
county_data<-get_acs(geography="county",STATE="PA",variables=paper_variables,year=2015, output="wide") 

#See the variables in the dataset
objects(county_data)

county_data<-subset(county_data,GEOID>42000) #Only keep information for PA (safety check)
county_data<-subset(county_data,GEOID<42999) #Only keep information for PA (safety check)

library(readxl) #With this library we import the Urban Rural Codes for 2013
ruralurbancodes2013 <- read_excel("PATH HERE/ruralurbancodes2013.xls")

ruralurbancodes2013$GEOID<-as.numeric(ruralurbancodes2013$FIPS) #Make FIPS a numeric variable

ruralurbancodes2013<-subset(ruralurbancodes2013,GEOID>42000) #Only keep information for PA
ruralurbancodes2013<-subset(ruralurbancodes2013,GEOID<42999) #Only keep information for PA
```

##Variables considered

The following variables will be considered. 

* Poverty Rate
* Percent non-Hispanic Black
* Percent Hispanic
* Percent of Adults without a High School Diploma
* Unemployment Rate
* Percent Female Headed Households
* Percent 65 years and over
* Gini Index of Income Inequality
* Poverty Segregation Index (Interaction Index between poor and non-poor)
* Percent of the population without health insurance
* Median Household Income
* Median House Value (for houses occupied by owner)

In the following line we will calculate the variables to be considered using R. 

```{r calculate variables}

attach(county_data) #Will make the data part of R, and then will send it back to the original database. 
#This does not requires us to write "paper_data" everytime in front of each variable. 
#Note that we are creating the variables inside "paper_data"

county_data$pov_percent<-B17001_002E/B17001_001E #Calculates the percent of the population below the poverty level

county_data$pct_black<-B03002_004E/B03002_001E*100 #Calculates the percent non-Hispanic black

county_data$pct_hispanic<-B03002_012E/B03002_001E*100

county_data$pct_under18years<-(B01001_003E+B01001_004E+B01001_005E+B01001_006E+B01001_027E+B01001_028E+B01001_029E+B01001_030E)/B01001_001E*100 #Calculates the percent of the population under 18 years (0-17 years old)
  
county_data$pct_65andolder<-(B01001_020E+B01001_021E+B01001_022E+B01001_023E+B01001_024E+B01001_025E+B01001_044E+B01001_045E+B01001_046E+
                            B01001_047E+B01001_048E+B01001_049E)/B01001_001E*100
#Calculates the percent of the population that is 65 years and older

county_data$pct_immobile_1yr<-(B07003_004E+B07003_007E)/B07003_001E*100 #Calculates percent immbole 1 year ago
  
county_data$percent_lessthanHS<-(B15003_002E+B15003_003E+B15003_004E+B15003_005E+B15003_006E+B15003_007E+B15003_008E+B15003_009E+B15003_010E+B15003_011E+B15003_012E+B15003_013E+B15003_014E+B15003_015E+B15003_016E)/B15003_001E*100
#Calculates the percent of the population over 25 years without a high school education 

county_data$pct_manufacturing<-C24050_004E/C24050_001E*100 #Population working in Manufacturing Sector 
  
county_data$pct_extractive<-C24050_002E/C24050_001E*100    #Population working in Agriculture, Forestry, Fishing, and Mining
  
county_data$pct_service<-C24050_029E/C24050_001E*100       #Population woking in the Service Sector
  
county_data$pct_government<-C24050_014E/C24050_001E*100    #Population working in Public Administration (Government)

county_data$pct_female_headed_hh<-(B25011_013E+B25011_037E)/B25011_001E*100 #Female Headed Households

county_data$pct_fem_employed<-(B12006_010E+B12006_021E+B12006_032E+B12006_043E+B12006_054E)/(B12006_008E+B12006_019E+B12006_030E+B12006_041E+B12006_052E)*100 #Percent of females employed

county_data$pct_unemployed<-(B17005_006E+B17005_011E+B17005_017E+B17005_022E)/B17005_001E*100 #Percent unemployed

county_data$gini<-B19083_001E

life_exp<-read.csv("PATH HERE/Life_Expectancy_PA_data.csv") #Read in the Life Expectancy File

life_exp$GEOID<-life_exp$FIPS

life_exp$GEOID<-as.character(life_exp$GEOID)

full_data<-merge(life_exp,county_data,by="GEOID") # Merge county data with life expectancy data into full_data

ruralurbancodes2013$GEOID<-as.character(ruralurbancodes2013$GEOID) #Make FIPS in RUCC file a character

ruralurbancodes2013$nonmetro<-ifelse(ruralurbancodes2013$RUCC_2013>3,1,0) #Value above 3 is non-metro (nonmetro=1), those below are metro (nonmetro=0)

full_data<-merge(full_data,ruralurbancodes2013,by="GEOID") #Include rural urban codes in full_data

table(full_data$nonmetro) #Cross tabulation of metro/non-metro

social_capital<-read_excel("PATH HERE/Social_Capital_2014.xlsx") #Reads in the social capital index file

social_capital<-subset(social_capital,FIPS>42000) #Keeps only PA variables
social_capital<-subset(social_capital,FIPS<42999) #Keeps only PA variables

social_capital$GEOID<-as.character(social_capital$FIPS) #Renames FIPS as GEOID for merge

full_data<-merge(full_data,social_capital,by="GEOID") #Merge social capital to the analytic dataset

full_data$medhhincome<-full_data$B25099_001E #Rename

full_data$medhousevalue<-full_data$B25077_001E #Rename

full_data$noinsurance<-(B27010_017E+B27010_033E+B27010_050E+B27010_066E)/B27010_001E*100 #No Health Insurance

density<-read.csv("PATH HERE/Land_Area.csv") #Land Area for Population Density

full_data<-merge(full_data,density,by="GEOID") #Merges the Land Area to full_data

model2<-lm(Diff_average_14~nonmetro+pov_percent+pct_black+pct_hispanic+pct_under18years+pct_65andolder+percent_lessthanHS+pct_manufacturing+pct_extractive+pct_service+pct_government+pct_unemployed+pct_female_headed_hh,data=full_data)

model3<-lm(Diff_average_14~pov_percent+pct_black+pct_hispanic+pct_under18years+pct_65andolder+percent_lessthanHS+pct_unemployed+pct_female_headed_hh,data=full_data)

model4<-lm(Life_exp_14~pov_percent+pct_black+pct_hispanic+pct_under18years+pct_65andolder+percent_lessthanHS+pct_unemployed+pct_female_headed_hh,data=full_data)

f<-ggplot(full_data,aes(pov_percent,Diff_average_14))
f+geom_jitter()

f<-ggplot(full_data,aes(percent_lessthanHS,Diff_average_14))
f+geom_jitter()


f+geom_jitter()

f<-ggplot(full_data,aes(gini,Diff_average_14))
f+geom_jitter()
```

#Correlations

```{r correlations}
z<-ggplot(full_data, aes(x=gini, y=Diff_average_14)) + 
  geom_point(color='blue') + 
  geom_smooth(method=lm, se=FALSE, fullrange=TRUE)

z+labs(title="Gini Index of Income Inequality and life expectancy differences from state level",subtitle="Average life expectancy for PA was 78.82 years",x="Poverty Rate",y="Difference in life expectancy (diff=county-state level)")

corr<-cor.test(x=full_data$gini,y=full_data$Diff_average_14,method='pearson')
corr

a<-ggplot(full_data, aes(x=pov_percent, y=Diff_average_14)) + 
  geom_point(color='blue') + 
  geom_smooth(method=lm, se=FALSE, fullrange=TRUE)

a+labs(title="Poverty rate and life expectancy differences from state level",subtitle="Average life expectancy for PA was 78.82 years",x="Poverty Rate",y="Difference in life expectancy (diff=county-state level)")

corr<-cor.test(x=full_data$pov_percent,y=full_data$Diff_average_14,method='pearson')
corr

aa<-ggplot(full_data, aes(x=pct_black, y=Diff_average_14)) + 
  geom_point(color='blue') + 
  geom_smooth(method=lm, se=FALSE, fullrange=TRUE)

aa+labs(title="Percent Black and life expectancy differences from state level",subtitle="Average life expectancy for PA was 78.82 years",x="Percent Black",y="Difference in life expectancy (diff=county-state level)")

corr<-cor.test(x=full_data$pct_black,y=full_data$Diff_average_14,method='pearson')
corr

dummy<-subset(full_data,pct_black<30)

aa<-ggplot(dummy, aes(x=pct_black, y=Diff_average_14)) + 
  geom_point(color='blue') + 
  geom_smooth(method=lm, se=FALSE, fullrange=TRUE)

aa+labs(title="Percent Black and life expectancy differences from state level",subtitle="Average life expectancy for PA was 78.82 years",x="Percent Black",y="Difference in life expectancy (diff=county-state level)",caption="Note: Removing outlier")

corr<-cor.test(x=dummy$pct_black,y=dummy$Diff_average_14,method='pearson')
corr

b<-ggplot(full_data, aes(x=sk2014, y=Diff_average_14)) + 
  geom_point(color='blue') + 
  geom_smooth(method=lm, se=FALSE, fullrange=TRUE)

b+labs(title="Social Capital and life expectancy differences from state level",subtitle="Average life expectancy for PA was 78.82 years",x="Social Capital Index",y="Difference in life expectancy (diff=county-state level)")

corr<-cor.test(x=full_data$sk2014,y=full_data$Diff_average_14,method='pearson')
corr

c<-ggplot(full_data, aes(x=log(medhhincome), y=Diff_average_14)) + 
  geom_point(color='blue') + 
  geom_smooth(method=lm, se=FALSE, fullrange=TRUE)

c+labs(title="Median Household Income and life expectancy differences from state level",subtitle="Average life expectancy for PA was 78.82 years",x="Median Household Income (log scale)",y="Difference in life expectancy (diff=county-state level)")

corr<-cor.test(x=log(full_data$medhhincome),y=full_data$Diff_average_14,method='pearson')
corr

d<-ggplot(full_data, aes(x=log(medhousevalue), y=Diff_average_14)) + 
  geom_point(color='blue') + 
  geom_smooth(method=lm, se=FALSE, fullrange=TRUE)

d+labs(title="Median House Value (Owner-occupied) and life expectancy differences from state level",subtitle="Average life expectancy for PA was 78.82 years",x="Median Household Value (log scale)",y="Difference in life expectancy (diff=county-state level)")

corr<-cor.test(x=log(full_data$medhousevalue),y=full_data$Diff_average_14,method='pearson')
corr
```


```{r segregation}
#Measuring as Interaction index
#get total and poor/nonpoor

pov_var<-c("B17001_001E","B17001_002E","B17001_031E")

tract<-get_acs(geography="tract",state ="42",variables=pov_var,year=2015, output="wide") 

tract$total<-tract$B17001_001E

tract$poor<-tract$B17001_002E

tract$nonpoor<-tract$B17001_031E

tract$cofips<-substr(tract$GEOID,1,5)

#County total

co_total<-tapply(tract$total, tract$cofips, sum)
co_total<-data.frame(cofips=names(unlist(co_total)), pop=unlist(co_total))

#POOR
co_poor<-tapply(tract$poor, tract$cofips, sum)
co_poor<-data.frame(cofips=names(unlist(co_poor)), pop=unlist(co_poor))

#non-poor
co_nonpoor<-tapply(tract$nonpoor, tract$cofips, sum)
co_nonpoor<-data.frame(cofips=names(unlist(co_nonpoor)), pop=unlist(co_nonpoor))

c_ppoor<-co_poor$pop/co_total$pop
c_pnonpoor<-co_nonpoor$pop/co_total$pop

#Poverty Segregation
pov_seg<-data.frame(cofips=co_total$cofips, co_total=co_total$pop, co_poor_total=co_poor$pop, co_nonpoor_total=co_nonpoor$pop, c_poor=c_ppoor, c_pnonpoor=c_pnonpoor)

merged<-merge(x=tract,y=pov_seg, by="cofips", all.x=T )

#The interaction index measures the exposure of minority group members to members of the majority group as the 
# minority-weighted average of the majority proportion of the population in each areal unit. The isolation index
# measures "the extent to which minority members are exposed only to one another," and is computed as the
# minority-weighted average of the minority proportion in each area.  

merged$int.wb<-(merged$poor/merged$co_poor_total * merged$nonpoor/merged$total)

int.wb.tr<-tapply(merged$int.wb, merged$cofips, sum, na.rm=T)

county_seg<-data.frame(GEOID=names(unlist(int.wb.tr)),seg=unlist(int.wb.tr))
```

```{r merge}
full_data<-merge(full_data,county_seg,by="GEOID") #Merge with Segregation Index

e<-ggplot(full_data, aes(x=seg, y=Diff_average_14)) + 
  geom_point(color='blue') + 
  geom_smooth(method=lm, se=FALSE, fullrange=TRUE)

e+labs(title="Poor/Non-Poor Segregation and life expectancy differences from state level",subtitle="Average life expectancy for PA was 78.82 years",x="Poverty Segregation",y="Difference in life expectancy (diff=county-state level)",caption="Interaction Index is used to measure segregation")

corr<-cor.test(x=log(full_data$seg),y=full_data$Diff_average_14,method='pearson')
corr

dummy2<-subset(full_data,seg>0.70)

ee<-ggplot(dummy2, aes(x=seg, y=Diff_average_14)) + 
  geom_point(color='blue') + 
  geom_smooth(method=lm, se=FALSE, fullrange=TRUE)

ee+labs(title="Poor/Non-Poor Segregation and life expectancy differences from state level",subtitle="Average life expectancy for PA was 78.82 years",x="Poverty Segregation",y="Difference in life expectancy (diff=county-state level)",caption="Interaction Index is used to measure segregation and removing outliers (low end)")

corr<-cor.test(x=log(dummy2$seg),y=dummy2$Diff_average_14,method='pearson')
corr

f<-ggplot(full_data, aes(x=noinsurance, y=Diff_average_14)) + 
  geom_point(color='blue') + 
  geom_smooth(method=lm, se=FALSE, fullrange=TRUE)

f+labs(title="No-Health Insurance and life expectancy differences from state level",subtitle="Average life expectancy for PA was 78.82 years",x="No Insurance Rate",y="Difference in life expectancy (diff=county-state level)")

corr<-cor.test(x=full_data$noinsurance,y=full_data$Diff_average_14,method='pearson')
corr

g<-ggplot(full_data, aes(x=pct_unemployed, y=Diff_average_14)) + 
  geom_point(color='blue') + 
  geom_smooth(method=lm, se=FALSE, fullrange=TRUE)


g+labs(title="Percent Unemployed and life expectancy differences from state level",subtitle="Average life expectancy for PA was 78.82 years",x="Percent Unemployed",y="Difference in life expectancy (diff=county-state level)")

corr<-cor.test(x=full_data$pct_unemployed,y=full_data$Diff_average_14,method='pearson')
corr

h<-ggplot(full_data, aes(x=pct_female_headed_hh, y=Diff_average_14)) + 
  geom_point(color='blue') + 
  geom_smooth(method=lm, se=FALSE, fullrange=TRUE)

h+labs(title="Female Headed Households and life expectancy differences from state level",subtitle="Average life expectancy for PA was 78.82 years",x="Percent Female Headed Households",y="Difference in life expectancy (diff=county-state level)")

corr<-cor.test(x=full_data$pct_female_headed_hh,y=full_data$Diff_average_14,method='pearson')
corr
  
i<-ggplot(full_data, aes(x=pct_65andolder, y=Diff_average_14)) + 
  geom_point(color='blue') + 
  geom_smooth(method=lm, se=FALSE, fullrange=TRUE)

i+labs(title="Percent of the Population 65 + and life expectancy differences from state level",subtitle="Average life expectancy for PA was 78.82 years",x="Percent of the Population 65 +",y="Difference in life expectancy (diff=county-state level)")

corr<-cor.test(x=full_data$pct_65andolder,y=full_data$Diff_average_14,method='pearson')
corr

j<-ggplot(full_data, aes(x=percent_lessthanHS, y=Diff_average_14)) + 
  geom_point(color='blue') + 
  geom_smooth(method=lm, se=FALSE, fullrange=TRUE)

j+labs(title="% Adult Population without High School Diploma and life expectancy differences from state level",subtitle="Average life expectancy for PA was 78.82 years",x="Percent of the Population 65 +",y="Difference in life expectancy (diff=county-state level)")


corr<-cor.test(x=full_data$percent_lessthanHS,y=full_data$Diff_average_14,method='pearson')
corr

k<-ggplot(full_data, aes(x=pct_black, y=Diff_average_14)) + 
  geom_point(color='blue') + 
  geom_smooth(method=lm, se=FALSE, fullrange=TRUE)

k+labs(title="% Non-Hispanic Black and life expectancy differences from state level",subtitle="Average life expectancy for PA was 78.82 years",x="Percent of the Population non-Hispanic Black",y="Difference in life expectancy (diff=county-state level)")

corr<-cor.test(x=full_data$pct_black,y=full_data$Diff_average_14,method='pearson')
corr

l<-ggplot(full_data, aes(x=pct_hispanic, y=Diff_average_14)) + 
  geom_point(color='blue') + 
  geom_smooth(method=lm, se=FALSE, fullrange=TRUE)

l+labs(title="% Hispanic and life expectancy differences from state level",subtitle="Average life expectancy for PA was 78.82 years",x="Percent of the Population Hispanic",y="Difference in life expectancy (diff=county-state level)")

corr<-cor.test(x=full_data$pct_hispanic,y=full_data$Diff_average_14,method='pearson')
corr

dummy4<-subset(full_data,pct_hispanic<15)

m<-ggplot(dummy4, aes(x=pct_hispanic, y=Diff_average_14)) + 
  geom_point(color='blue') + 
  geom_smooth(method=lm, se=FALSE, fullrange=TRUE)

m+labs(title="% Hispanic and life expectancy differences from state level",subtitle="Average life expectancy for PA was 78.82 years",x="Percent of the Population Hispanic",y="Difference in life expectancy (diff=county-state level)",caption="Removing outliers for % Hispanic population")

corr<-cor.test(x=dummy4$pct_hispanic,y=dummy4$Diff_average_14,method='pearson')
corr

lm(full_data$Diff_average_14~full_data$nonmetro)

t.test(full_data$Diff_average_14~full_data$nonmetro)

aov(formula =full_data$Diff_average_14~full_data$nonmetro)

n<-ggplot(full_data, aes(x=Density, y=Diff_average_14)) + 
  geom_point(color='blue') + 
  geom_smooth(method=lm, se=FALSE, fullrange=TRUE)

n+labs(title="Population Density and life expectancy differences from state level",subtitle="Average life expectancy for PA was 78.82 years",x="Population Density",y="Difference in life expectancy (diff=county-state level)")

corr<-cor.test(x=full_data$Density,y=full_data$Diff_average_14,method='pearson')
corr

dummy5<-subset(full_data,Density<3000)

o<-ggplot(dummy5, aes(x=Density, y=Diff_average_14)) + 
  geom_point(color='blue') + 
  geom_smooth(method=lm, se=FALSE, fullrange=TRUE)

o+labs(title="Population Density and life expectancy differences from state level",subtitle="Average life expectancy for PA was 78.82 years",x="Population Density",y="Difference in life expectancy (diff=county-state level)",caption="Removing two outliers")

corr<-cor.test(x=dummy5$Density,y=dummy5$Diff_average_14,method='pearson')
corr
```

##Maps
The following code produces maps

```{r map}
map_pa<-get_decennial(geography = "county", state="PA",year=2010 ,variables="P0130001",output ="wide",geometry = TRUE) #Requests the map with total population

map_pa %>%
  ggplot()+  geom_sf() #Produces a map 

map_pa2<-left_join(x = map_pa, y = full_data, by = "GEOID")

map_pa2 %>%
  ggplot(aes(fill = Diff_average_14))+  geom_sf()+labs(title = "County life-expectancy difference to state average", subtitle = "Darker represents a lower life expectancy than average", caption="Produced by PSU-Administrative Data Accelerator",fill="")

map_pa2 %>%
  ggplot(aes(fill =Life_exp_14))+  geom_sf()+labs(title = "County level life-expectancy", subtitle = "Darker represents a lower life expectancy than average", caption="Produced by PSU-Administrative Data Accelerator",fill="")

```
